Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-07 Manuscript/manuscriptPGPEsuppl.tex
Update manuscript

\section{Lipid headgroups in mixtures with PC and PE or PG lipids}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{P-N vector angles from Berger-OPLS and GROMOS-CKP simulations are yet to be analyzed.}
\end{figure*}

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
by the tested CHARMM36 and
Berger-OPLS force fields, although CHARMM36 slightly overestimates the changes 
The good performance of Berger-OPLS simulations is notable because the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/Höltje force field in our previous work \cite{botan15}.
%\todo{This is text by P. Fuchs, copied from the blog. \\
%Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
%- pure POPC \\
%CHARMM36: 0.624 \\
%Berger : 0.649 \\
%- POPC/POPE 50:50 \\
%CHARMM36 : POPC 0.609, POPE 0.557 \\
%Berger-hacked: POPC 0.637, POPE 0.632 \\
%----- \\
%One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
%In contrast, the values for Berger are not that changed. 
%The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
%-------\\
%The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
%which supports the more condensed PE bilayer.
%This is interesting, but not exactly the core message of the manuscript.
%Maybe we should mention this very briefly? For example, we could just report the areas per lipid (without distinguishing PC and PE)
%and mention the difference between CHARMM36 and Berger. I have opened an issue for this: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/7
%}

None of the force fields fully reproduces the PC headgroup order parameter response to the increasing amount of
PG, which may be related to the counterion binding affinity (see also the next section) \cite{antila19}.
In all force fields except Slipids, the order parameters of different
hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
\todo{Maybe we should figure out what is the reason for this? \\
  Maybe we should analyze the P-N vector angle from different simulations? \\
  https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/10}.

For $\beta$-carbon order parameter in PG headgroup, experiments report
mild increase \cite{macdonald87} or no change~\cite{borle85} upon addition 
of PC lipids (Fig. \ref{HGorderparametersPGvsPCchange}). 
Simulations with all the tested force fields give only very small changes also for
the $\alpha$-carbon order parameter (Figs. \ref{HGorderparametersPGvsPC} and \ref{HGorderparametersPGvsPCchange}). 
Therefore, the simulations are generally in line with experiments, suggesting that the
interactions with PC do not essentially effect the PG headgroup structure.
This suggests that the interactions between PG and PC headgroups are captured better in
simulations than for PS headgroup, where
all the force fieds significantly overestimated the structural response of PS headgroup
to the interactions with PC lipids \cite{antila18}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPCchange.eps}
  \caption{\label{HGorderparametersPGvsPCchange}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
  \todo{P-N angles from Slipids and GROMOS-CKP yet to be calculated.}
\end{figure}

\clearpage
\subsection{Sodium binding to PE and PG lipid bilayers}
Sodium binding affinity to PE lipids has not been measured experimentally,
but large differences to PC would be surprising. 
In simulations, the sodium binding affinity to POPE depends on the used force field (Fig. \ref{NAdensPE}),
but lesser extend than reported previously for PC~\cite{catte16}.
\todo{This will be finished once we have all the simulation details and Lipid17 simulations with correct dihedrals from issue https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12},
Because some simulation and ion parameters are not identical with the previous work~\cite{catte16},
we compare POPE results to the POPC simulations ran with identical parameters (Fig. \ref{NAdensPC}).
In Lipid17 with the strongest sodium binding affinity to POPE,
the binding affinity is approximately similar to POPC.
Slipids and CHARMM36 exhibit slightly, and
GROMOS-CKP subtantially weaker binding to POPE than to POPC.
%The Lipid17 exhibits stronger sodium binding affinity
%to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
%Experimental data for sodium binding to PE lipids is not available, while
%simulations in agreement with electrometer data from NMR suggests that
%sodium binding to POPC is weaker than in any of the simulations
%here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}.
Assuming that the binding
to POPE would be similar than to POPC, the sodium binding affinity to POPE
is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
but substantially overestimated in Lipid17 simulation.
%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%previous Amber lipid version, Lipid14.
%\todo{The potential differences in the used ion models should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

Simulations with PG lipids give similar dependence on force field as observed in POPE simulations:
Lipid 17 simulations with Dang ion parameters exhibits stronger counter-ion binding affinity to pure POPG bilayer
than CHARMM36, Slipids, and GROMOS-CKP simulations, which are roughly similar (Fig. \ref{CIdensPG}).
Lipid17 also exhibits less increase in POPC headgroup
%$\beta$ carbon
order parameters upon addition of POPG than other simulations (Fig. \ref{HGorderparametersPCvsPEPG}),
and lower area per molecule (59.5 {\AA}$^2$) than in experiments (66.1 {\AA}$^2$).
In our previous study about PS lipids \cite{antila18}, such behaviour was related to 
%the results suggest that
the overestimated counterions binding and shielding the electrostatic repulsion between PG headgroups
in bilayers.% simulated with Lipid17 parameters.
%the underestimated increase of PC
%headgroup order parameters upon addition of negaticely charged lipids
%was related
%to the overestimated counterion binding affinity, which would overcompensates the effect of negative charge.
%Similar correlation is observed here in %other than GROMOS-CKP simulation:
%%Lipid 17 simulations, which exhibits stronger counter-ion binding affinity to pure POPG bilayer (Fig. \ref{CIdensPG})
%
Even though the area per lipid in CHARMM36, Slipids, and GROMOS-CKP simulations is in good agreement
with experiments (Fig. \ref{CIdensPG}), the experimental increase in 
%In GROMOS-CKP simulations, however, the $\beta$-carbon order parameter of
POPC headgroup order parameters upon addition of POPG are not fully reproduced (Fig. \ref{HGorderparametersPCvsPEPG}).
%more than in CHARMM36 and Slipids simulations even though the binding affinities in these
%simulations are similar. On the other hand, the responses of different hydrogens attached in $\alpha$-carbons
%are different in all simulations except in Slipids, most sever in Lipid17 and GROMOS-CKP where other in increases
%and other decreases \todo{Can we figure out the reason for this?}.
Therefore we conclude that the counter-ion binding affinity is overestimated in Lipid17 simulations,
% with the strongest affinity among the tested models,
while the other simulations are more realistic, but slight overbinding cannot be excluded.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities and area per lipids from simulations with PG lipids.
    Experimental area for POPG at 303 K is 66.1 {\AA}$^2$ and 67 {\AA}$^2$ for DPPC at 323 K \cite{pan12b}.
  }
\end{figure}

\section{Calcium density distributions}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG1PC1.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (1:1) mixtures with different force fields.
  }
\end{figure}

\section{Changes of PG headroup order parameters upon addition of PC}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPC.eps}
  \caption{\label{HGorderparametersPGvsPC}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
\end{figure}

\section{Sodium binding to POPC simulations}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPC.eps}
  \caption{\label{NAdensPC}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PC lipids.
  }
  \todo{Discussion about differences to the NMRlipids II to be discussed once we have the details on ions models.}
\end{figure}

\section{Calcium binding to POPC:POPG (4:1) mixtures}
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC4.eps}
  \caption{\label{changesWITHCaClPG1PC4}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (4:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{macdonald87} and simulations.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
  }
  \todo{Lipid17ecc data to be analyzed and added.}
\end{figure*}

>>>>

\section{PC headgroup in mixtures with PE or PG lipids}
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of POPC headgroup order parameters with increasing amount of POPE (left) and POPG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure*}

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
%
%
%\todo{This is text by P. Fuchs, copied from the blog. \\
%Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
%- pure POPC \\
%CHARMM36: 0.624 \\
%Berger : 0.649 \\
%- POPC/POPE 50:50 \\
%CHARMM36 : POPC 0.609, POPE 0.557 \\
%Berger-hacked: POPC 0.637, POPE 0.632 \\
%----- \\
%One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
%In contrast, the values for Berger are not that changed. 
%The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
%-------\\
%The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
%which supports the more condensed PE bilayer.
%This is interesting, but not exactly the core message of the manuscript.
%Maybe we should mention this very briefly? For example, we could just report the areas per lipid (without distinguishing PC and PE)
%and mention the difference between CHARMM36 and Berger. I have opened an issue for this: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/7
%}
%
%
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neturalizes the effect of added negative charge \cite{antila19}.
%In all force fields except Slipids, the order parameters of different
%hydrogens attached to the $\alpha$-carbon are responsing differently when mixed with PE or PG lipids
%\todo{Maybe we should figure out what is the reason for this? \\
%  Maybe we should analyze the P-N vector angle from different simulations? \\
%  https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/10}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations in here is surprising because
headgroup conformational enemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

\clearpage
\section{PG headgroup in mixtures with PC lipids}
Changes in other than PC lipid headgroup with changing membrane composition are less
extensively characterized in the literature.
The $\beta$-carbon order parameter in PG headgroup increases
mildly \cite{macdonald87} or is unchanged~\cite{borle85} upon increasing amount
of PC lipids (Fig. \ref{HGorderparametersPGvsPCchange}), but experimental data from
$\alpha$-carbon is not available. Also the tested force fields predict very small changes
for the $\beta$-carbon order parameter, while the P-N vector tilt and its response to the
increased amount of PC varies significantly between force fields in figure \ref{HGorderparametersPGvsPCchange}. 
Therefore, more experimental data and more accurate force fields are still required
to resolve the PG conformational ensembles in mixtures with other lipids.
%the simulations are generally in line with experiments, suggesting that the
%interactions with PC do not essentially effect the PG headgroup structure.
%This suggests that the interactions between PG and PC headgroups are captured better in
%simulations than for PS headgroup, where
%all the force fieds significantly overestimated the structural response of PS headgroup
%to the interactions with PC lipids \cite{antila18}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPCchange.eps}
  \caption{\label{HGorderparametersPGvsPCchange}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
\end{figure}

%
%\clearpage
%\subsection{Sodium binding to PE and PG lipid bilayers}
%Sodium binding affinity to PE lipids has not been measured experimentally,
%but large differences to PC would be surprising. 
%In simulations, the sodium binding affinity to POPE depends on the used force field (Fig. \ref{NAdensPE}),
%but lesser extend than reported previously for PC~\cite{catte16}.
%\todo{This will be finished once we have all the simulation details and Lipid17 simulations with correct dihedrals from issue https://github.com/NM%RLipids/NMRlipidsIVPEandPG/issues/12},
%Because some simulation and ion parameters are not identical with the previous work~\cite{catte16},
%we compare POPE results to the POPC simulations ran with identical parameters (Fig. \ref{NAdensPC}).
%In Lipid17 with the strongest sodium binding affinity to POPE,
%the binding affinity is approximately similar to POPC.
%Slipids and CHARMM36 exhibit slightly, and
%GROMOS-CKP subtantially weaker binding to POPE than to POPC.
%%The Lipid17 exhibits stronger sodium binding affinity
%%to POPE bilayers than Slipids, CHARMM36 and GROMOS-CKP (Fig. \ref{NAdensPE}).
%%Experimental data for sodium binding to PE lipids is not available, while
%%simulations in agreement with electrometer data from NMR suggests that
%%sodium binding to POPC is weaker than in any of the simulations
%%here (Fig. \ref{NAdensPC})~\cite{catte16,melcr18}.
%Assuming that the binding
%to POPE would be similar than to POPC, the sodium binding affinity to POPE
%is potentially realistic in CHARMM36, Slipids, and GROMOS-CKP simulations here,
%but substantially overestimated in Lipid17 simulation.
%%On the other hand, the sodium binding affinity is stronger to POPE in Lipid17 simulations than to POPC in simulations with
%%previous Amber lipid version, Lipid14.
%%\todo{The potential differences in the used ion models should be checked}.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
%  \caption{\label{NAdensPE}
%    Sodium (solid line) and choride ion density profiles along membrane normal
%    from different simulations with PE lipids.
%  }
%\end{figure}

%Simulations with PG lipids give similar dependence on force field as observed in POPE simulations:
%Lipid 17 simulations with Dang ion parameters exhibits stronger counter-ion binding affinity to pure POPG bilayer
%than CHARMM36, Slipids, and GROMOS-CKP simulations, which are roughly similar (Fig. \ref{CIdensPG}).
%Lipid17 also exhibits less increase in POPC headgroup
%%$\beta$ carbon
%order parameters upon addition of POPG than other simulations (Fig. \ref{HGorderparametersPCvsPEPG}),
%and lower area per molecule (59.5 {\AA}$^2$) than in experiments (66.1 {\AA}$^2$).
%In our previous study about PS lipids \cite{antila18}, such behaviour was related to 
%%the results suggest that
%the overestimated counterions binding and shielding the electrostatic repulsion between PG headgroups
%in bilayers.% simulated with Lipid17 parameters.
%%the underestimated increase of PC
%%headgroup order parameters upon addition of negaticely charged lipids
%%was related
%%to the overestimated counterion binding affinity, which would overcompensates the effect of negative charge.
%%Similar correlation is observed here in %other than GROMOS-CKP simulation:
%%%Lipid 17 simulations, which exhibits stronger counter-ion binding affinity to pure POPG bilayer (Fig. \ref{CIdensPG})
%%
%Even though the area per lipid in CHARMM36, Slipids, and GROMOS-CKP simulations is in good agreement
%with experiments (Fig. \ref{CIdensPG}), the experimental increase in 
%%In GROMOS-CKP simulations, however, the $\beta$-carbon order parameter of
%POPC headgroup order parameters upon addition of POPG are not fully reproduced (Fig. \ref{HGorderparametersPCvsPEPG}).
%%more than in CHARMM36 and Slipids simulations even though the binding affinities in these
%%simulations are similar. On the other hand, the responses of different hydrogens attached in $\alpha$-carbons
%%are different in all simulations except in Slipids, most sever in Lipid17 and GROMOS-CKP where other in increases
%%and other decreases \todo{Can we figure out the reason for this?}.
%Therefore we conclude that the counter-ion binding affinity is overestimated in Lipid17 simulations,
%% with the strongest affinity among the tested models,
%while the other simulations are more realistic, but slight overbinding cannot be excluded.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
%  \caption{\label{CIdensPG}
%    Counterion densities and area per lipids from simulations with PG lipids.
%    Experimental area for POPG at 303 K is 66.1 {\AA}$^2$ and 67 {\AA}$^2$ for DPPC at 323 K \cite{pan12b}.
%  }
%\end{figure}

%\clearpage
%
%\section{Changes of PG headroup order parameters upon addition of PC}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPC.eps}
%  \caption{\label{HGorderparametersPGvsPC}
%    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
%    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
%  }
%\end{figure}

%\clearpage

%\section{Sodium binding to POPC simulations}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/NAdensPC.eps}
%  \caption{\label{NAdensPC}
%    Sodium (solid line) and choride ion density profiles along membrane normal
%    from different simulations with PC lipids.
%  }
%  \todo{Discussion about differences to the NMRlipids II to be discussed once we have the details on ions models.}
%\end{figure}

\clearpage
%\section{Calcium density distributions}
\section{Calcium binding to POPC:POPG mixtures}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CAdensPG1PC1.eps}
  \caption{\label{CAdensPG}
    Calcium ion density profiles along membrane normal from simulations of POPC:POPG (1:1) mixtures with different force fields.
    The changes in the order parameters upon addition of CaCl$_2$ are compared with experiments in figure \ref{changesWITHCaClPG} in the main text.
  }
\end{figure}

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC4.eps}
  \caption{\label{changesWITHCaClPG1PC4}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (4:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{macdonald87} and simulations.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
  }
\end{figure*}
